An investigation into the contentious phylogenetic results of the *Mixta* species

### 2021-01-27 Manuscript/Manuscript_Paper.Rmd
Update Manuscript_Paper.Rmd

## Distance Matrices
After alignments, model testing was done for each gene through `MEGA-CC` [@Kumar2012] for both nucleotide and amino acid sequences. `MEGA-CC` is called in a command-line terminal along with a text file listing the file pathways to all 799 genes and a MAO file detailing the analysis preferences. Default parameters were used for model testing. The output was a comma-separated values (CSV) file for each gene. These CSV files were read into `R`, and the model for phylogenetic analysis was extracted based on the lowest Bayesian Information Criterion (BIC). Twelve and twenty-three unique models were used for nucleotide (Table 2) and amino acid sequences (Table 3) were needed to create phylogenetic gene trees.

@@

## Genetic Distance
### Closest Relative from Genetic Distances
Genetic distance gives a numeric estimate of evolutionary change between species. The species that has the shortest genetic distance to a species of interest is considered the closest relative. For the following analysis, model testing was done first to estimate the best model for each gene. Next, individual phylogenetic gene trees were created, and it was from these that distance matrices were extracted. *P. septica* was the closest relative to 66.3% and 66.5% of the *M. calida* and *M. gaviniae* genes, respectively, for the nucleotide sequences (Figure 1). In comparison, only 42.7% and 42.8% of *M. calida* and *M. gaviniae* genes, respectively, are closely related to *P. septica* for the amino acid sequences (Figure 1), but this is still the majority. These results suggest that *P. septica* is the closest relative to *Mixta*, followed by *P. agglomerans*. In contrast, the *Mixta* genes are more closely related to the other species in the study than *P. septica* in the amino acid analysis (Figure 1). For instance, the percentage of genes related to *P. agglomerans* increased by 8.4 and 8.5% for *M. calida* and *M. gaviniae*, respectively (Figure 1). This may be due to a loss of phylogenetic signal when using amino acid sequences caused by codon degeneracy and having a third of available sites.

@@

```{r Figure2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 2. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species when using the recommended (best) model from model testing vs the restricted model (best available) options available for genetic distance analysis in MEGAX for nucleotide sequences. The species with the shortest genetic distance from both of the *Mixta* species is considered the closest relative. Evolutionary patterns may vary by gene; therefore, it is appropriate to conduct model testing and use the model that explains the most variation. Nevertheless, this model may not be available in software programs or analysis methods within a software program. Pink bars represent the percentage of *Mixta* genes most closely related to non-*Mixta* species when using the recommended model. Orange bars represent the percentage of *Mixta* genes most closely related to any non-*Mixta* species when using the best available genetic distance model analysis in MEGAX. GTR and HKY models and invariant sites parameter were not available for this analysis. The lighter shades represent *M. calida*, and the darker shades represent *M. gaviniae*.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure2_FirstRelativeBestvsAvailableModelNT.csv", 
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", 
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("PD_calida", "PD_gaviniae", "GD_calida", "GD_gaviniae"), ordered = TRUE))

@@

Although genetic distance estimation in `MEGAX` did not offer all of the models and parameters as suggested by model testing, all of the models and parameters were available for phylogenetic tree analysis and genetic distance can be exported from these trees. However, this has to be done manually by opening the tree file in `MEGAX` and exporting the distance matrix as either an Excel or CSV file. Therefore, it is not feasible when hundreds or thousands of genetic distance matrices are needed.

@@

Fewer genes were closely related to *P. septica* when analyzing the amino acid sequences (Figure 4). Instead, more genes were closely related to *P. agglomerans*, the *Erwinia* species, the *Tatumella* species, and *E. cloacae* (Figure 4). Nevertheless, no apparent patterns arose in the closest relatives to the *Mixta* genes (Figure 4). The *Mixta* genes that have changed their allegiance between the nucleotide and amino acid analyses may result from the redundancy of the genetic code.

@@

```{r Figure5, echo = FALSE, message = FALSE, warning=FALSE, fig.cap="Figure 5. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species using percent site identities. The species with the highest percent identity to each of the *Mixta* species is considered to be most similar taxon for a single gene. Blue bars are the results for the nucleotide sequences where the light blue bars represent *M. calida* sequences and the dark blue bars represent *M. gaviniae* sequences. Green bars are the results for the amino acid sequences where the light green bars represent *M. calida* sequences and the dark green bars represent *M. gaviniae* sequences.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure5_FirstRelativeMixtaNTAAIdentity.csv", 
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", 
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("calida_NT", "gaviniae_NT", "calida_AA", "gaviniae_AA"), ordered = TRUE))

@@

In contrast, only 47.8 and 47.6% of *M. calida* and *M. gaviniae* sequences were most similar to *P. septica* for the amino acid sequences (Figure 5). Instead, the genes are most similar to either *P. agglomerans* or one of the two *Erwinia* species (Figure 5). Additionally, 3.1 and 2.5% more genes were most similar to the *E. cloacae* sequences for *M. calida* and *M. gaviniae*, respectively (Figure 5). In the identity analyses, amino acids that are functionally equivalent were not considered. This may explain the large differences between the nucleotide and amino acid identities for which species was most similar to the *Mixta* sequences. Additionally, at every site for a nucleotide sequence, each base has a 1:3 ratio of being identical to the base at the corresponding location of the other sequence whereas there is only a 1:19 ratio of an amino acid being identical to the amino acid at the same position of the other sequence. Thus, it is far more likely for a nucleotide at a random position to be identical the nucleotide at the same position of the other sequence. This may lead to problems when only amino acid sequences are considered, especially if functional equivalence is not considered.

Although the numbers of genes that are most similar to *P. septica* are similar between the two *Mixta* species for both nucleotide and amino acid sequences (Figure 5), there were many genes in which the same species was the not the most similar taxa to both *Mixta* species. There were 86 genes for the nucleotide sequences and 147 genes for the amino acid sequences. However, of the 86 genes for nucleotide sequences, it was the case for 52% that one of the *Mixta* sequences was most similar to one representative of a genus and the other *Mixta* sequences was most similar to the other species of the same genus. This was also the case for 62% of the 147 amino acid genes.

@@

### Spatial Patterns of Most Similar Taxa
Similar to what was done with the closest relatives in the genetic distance analysis, spatial patterns of the most similar taxa to the *M. calida* sequences based on site identity were investigated for the nucleotide sequences (Figure 6) and the amino acid sequences (Figure 7). Once again, no discernible patterns arose when spatial position on the *M. calida* genome are considered when either the nucleotide or amino acid sequences are analyzed (Figure 6 and 7). However, majority of the nucleotide genes that were most similar to the two *Tatumella* species and to *E. amylovora* occur in the first half of the genome (Figure 6).

>>>>

## Distance Matrices
After alignments, model testing was done for each gene through `MEGA-CC` [@Kumar2012] for nucleotide and amino acid sequences. `MEGA-CC` is called in a command-line terminal along with a text file listing the file pathways to the 799 genes and a MAO file detailing the analysis preferences. Default parameters were used for model testing. The output was a comma-separated values (CSV) file for each gene. These CSV files were read into `R`, and the model for phylogenetic analysis was extracted based on the lowest Bayesian Information Criterion (BIC). Twelve and twenty-three unique models were used for nucleotide (Table 2) and amino acid sequences (Table 3) were needed to create phylogenetic gene trees.

@@

## Genetic Distance
### Closest Relative from Genetic Distances
Genetic distance gives a numeric estimate of evolutionary change between species. The species that has the shortest genetic distance to a species of interest is considered the closest relative. For the following analysis, model testing was done first to estimate the best model for each gene. Next, individual phylogenetic gene trees were created, and it was from these that distance matrices were extracted. *P. septica* was the closest relative to 66.3% and 66.5% of the *M. calida* and *M. gaviniae* genes, respectively, for the nucleotide sequences (Figure 1). In comparison, only 42.7% and 42.8% of *M. calida* and *M. gaviniae* genes, respectively, are closely related to *P. septica* for the amino acid sequences (Figure 1), but this is still the majority. These results suggest that *P. septica* is the closest relative to *Mixta*, followed by *P. agglomerans*. In contrast, the *Mixta* genes are more closely related to the other species in the study than *P. septica* in the amino acid analysis (Figure 1). For instance, the percentage of genes related to *P. agglomerans* increased by 8.4 and 8.5% for *M. calida* and *M. gaviniae*, respectively (Figure 1). This difference may be due to a loss of phylogenetic signal when using amino acid sequences caused by codon degeneracy and having a third of available sites.

@@

```{r Figure2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 2. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species when using the recommended (best) model from model testing vs the restricted model (best available) options available for genetic distance analysis in MEGAX for nucleotide sequences. The species with the shortest genetic distance from both of the *Mixta* species is considered the closest relative. Evolutionary patterns may vary by gene; therefore, it is appropriate to conduct model testing and use the model that explains the most variation. Nevertheless, this model may not be available in software programs or analysis methods within a software program. Pink bars represent the percentage of *Mixta* genes most closely related to non-*Mixta* species when using the recommended model. Orange bars represent the percentage of *Mixta* genes most closely related to non-*Mixta* species when using the best available genetic distance model analysis in MEGAX. GTR and HKY models and invariant sites parameter were not available for this analysis. The lighter shades represent *M. calida*, and the darker shades represent *M. gaviniae*.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure2_FirstRelativeBestvsAvailableModelNT.csv", 
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", 
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("PD_calida", "PD_gaviniae", "GD_calida", "GD_gaviniae"), ordered = TRUE))

@@

Although genetic distance estimation in `MEGAX` did not offer all of the models and parameters suggested by model testing, all of the models and parameters were available for phylogenetic tree analysis and genetic distance can be exported from these trees. However, this has to be done manually by opening the tree file in `MEGAX` and exporting the distance matrix as either an Excel or CSV file. Therefore, it is not feasible when hundreds or thousands of genetic distance matrices are needed.

@@

Fewer genes were closely related to *P. septica* when analyzing the amino acid sequences (Figure 4). Instead, more genes were closely related to *P. agglomerans*, the *Erwinia* species, the *Tatumella* species, and *E. cloacae* (Figure 4). Nevertheless, no apparent patterns arose in the closest relatives to the *Mixta* genes (Figure 4). The *Mixta* genes that have changed their allegiance between the nucleotide and amino acid analyses may result from the genetic code's redundancy.

@@

```{r Figure5, echo = FALSE, message = FALSE, warning=FALSE, fig.cap="Figure 5. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species using percent site identities. The species with the highest percent identity to each of the *Mixta* species is considered the most similar taxon for a single gene. Blue bars represent the nucleotide sequences' results; the light blue bars represent *M. calida* sequences, and the dark blue bars represent *M. gaviniae* sequences. Green bars represent the amino acid sequences' results; the light green bars represent *M. calida* sequences, and the dark green bars represent *M. gaviniae* sequences.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure5_FirstRelativeMixtaNTAAIdentity.csv", 
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", 
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("calida_NT", "gaviniae_NT", "calida_AA", "gaviniae_AA"), ordered = TRUE))

@@

In contrast, only 47.8 and 47.6% of *M. calida* and *M. gaviniae* sequences were most similar to *P. septica* for the amino acid sequences (Figure 5). Instead, the genes are most similar to either *P. agglomerans* or one of the two *Erwinia* species (Figure 5). Additionally, 3.1 and 2.5% more genes were most similar to the *E. cloacae* sequences for *M. calida* and *M. gaviniae*, respectively (Figure 5). In the identity analyses, functionally equivalent amino acids were not considered. This may explain the large differences between the nucleotide and amino acid identities for which species was most similar to the *Mixta* sequences. Furthermore, every site in a nucleotide sequence has a 1:3 chance of being identical to the site at the corresponding location of a homologous gene. In contrast, there is only a 1:19 chance for amino acids. Thus, two nucleotides at the same position on two homologues are more likely to be identical. This may lead to problems when only amino acid sequences are considered, especially if functional equivalence is not considered.

The two *Mixta* species have a comparable number of genes that are most similar to *P. septica* for both nucleotide and amino acid sequences (Figure 5). Nevertheless, many genes in which the two *Mixta* species did not share the same closest relative. For the nucleotide sequences, 10.8% of the *Mixta* genes did not share the same species, and 5.6% did not share the same genus for the closest relative. For the amino acid sequences, 18.3% did not share the same species, and 11.4% did not share the same genus.

Percent identity analyses provide only the similarity of two sequences. They do not estimate evolutionary history nor penalize changes in the sequences. However, since members of a genus should be genetically similar, the percent identity analyses lend support to two taxa's relatedness.

@@

### Spatial Patterns of Most Similar Taxa
**HERE!!**

Similar to what was done with the closest relatives in the genetic distance analysis, spatial patterns of the most similar taxa to the *M. calida* sequences based on site identity were investigated for the nucleotide sequences (Figure 6) and the amino acid sequences (Figure 7). Once again, no discernible patterns arose when spatial position on the *M. calida* genome are considered when either the nucleotide or amino acid sequences are analyzed (Figure 6 and 7). However, majority of the nucleotide genes that were most similar to the two *Tatumella* species and to *E. amylovora* occur in the first half of the genome (Figure 6).
